---
title: "Analysis of Variance (ANOVA)"
subtitle: "IS381 - Statistics and Probability with R"
author: Jason Bryer, Ph.D.
date: "November 17, 2025"
output:
  xaringan::moon_reader:
    css: ["assets/mtheme_max.css", "assets/fonts_mtheme_max.css"]
    lib_dir: libs
    seal: false
    nature:
      highlightStyle: solarized-light
      highlightLanguage: R
      highlightLines: true
      countIncrementalSlides: false
      ratio: "16:9"
      navigation:
        scroll: false
    includes:
      in_header: [assets/header.html]
      after_body: [assets/insert-logo.html]
editor_options: 
  chunk_output_type: console
---

```{r setup, include = FALSE}
# Cartoons from https://github.com/allisonhorst/stats-illustrations
# dplyr based upon https://allisonhorst.shinyapps.io/dplyr-learnr/#section-welcome

source('../config.R')
options(width = 120)
```

class: center, middle, inverse, title-slide

# `r metadata$title`
## `r metadata$subtitle`
### `r metadata$author`
### `r metadata$date`


---
# One Minute Paper Results

```{r, echo=FALSE, cache=FALSE}
library(googlesheets4)
omp <- read_sheet(one_minute_paper_results)
omp <- omp %>% dplyr::filter(Topic == 'Inference for Categorical Data (Chapter 6)')
```

.pull-left[
**What was the most important thing you learned during this class?**
```{r, echo=FALSE, fig.height=9}
ompWordCloud(omp$`What was the most important thing you learned during this class?`)
```
]
.pull-right[
**What important question remains unanswered for you?**
```{r, echo=FALSE, fig.height=9}
ompWordCloud(omp$`What important question remains unanswered for you?`)
```
]


```{r, echo=FALSE, results='hide', message=FALSE, warning=FALSE}
require(ggplot2)
require(gdata)
require(psych)
require(granova)
require(granovaGG)
require(lattice)
library(VisualStats)
library(ggbeeswarm)

data(singer)
data(rat)
data(hand_washing, package = 'VisualStats')
```
]



---
class: font90
# Analysis of Variance (ANOVA)

The goal of ANOVA is to test whether there is a discernible difference between the means of several groups.

**Hand Washing Example**

Is there a difference between washing hands with:  water only, regular soap, antibacterial soap (ABS), and antibacterial spray (AS)?

* Each tested with 8 replications
* Treatments randomly assigned

For ANOVA:

* The means all differ.
* Is this just natural variability?
* Null hypothesis:  All the means are the same.
* Alternative hypothesis:  The means are not all the same.

Source: De Veaux, R.D., Velleman, P.F., & Bock, D.E. (2014). *Intro Stats, 4th Ed.* Pearson.

---
# Boxplot

```{r, fig.height=6}
ggplot(hand_washing, aes(x = Method, y = Bacterial_Counts)) +  geom_boxplot() +
    geom_beeswarm(aes(color = Method)) + theme(legend.position = 'none')
```

---
class: font90
# Descriptive Statistics

```{r}
desc <- psych::describeBy(hand_washing$Bacterial_Counts, group = hand_washing$Method, mat = TRUE, skew = FALSE)
names(desc)[2] <- 'Method' # Rename the grouping column
desc$Var <- desc$sd^2 # We will need the variance latter, so calculate it here
desc
```

.pull-left[
```{r}
( k <- length(unique(hand_washing$Method)) )
( n <- nrow(hand_washing) )
```
]
.pull-right[
```{r}
( grand_mean <- mean(hand_washing$Bacterial_Counts) )
( grand_var <- var(hand_washing$Bacterial_Counts) )
( pooled_var <- mean(desc$Var) )
```
]

---
# Contrasts

A contrast is a linear combination of two or more factor level means with coefficients that sum to zero.

```{r}
desc$contrast <- (desc$mean - mean(desc$mean))
mean(desc$contrast) # Should be 0!
desc
```

---
# Plotting using contrasts

```{r, echo=FALSE}
anova_vis(Y = hand_washing$Bacterial_Counts, 
          group = hand_washing$Method,
          plot_boxplot = TRUE,
          plot_group_variances = FALSE,
          plot_group_sd = FALSE,
          plot_ms_within = FALSE,
          plot_ms_between = FALSE,
          plot_unit_line = FALSE,
          plot_grand_mean = FALSE,
          plot_sd_line = FALSE,
          plot_pooled_sd = FALSE,
          plot_between_group_variances = FALSE,
          ylab = 'Bacterial Counts'
)
```



---
# Grade Mean and Unit Line (slope = 1, intercept = $\bar{x}$)

```{r, echo = FALSE}
anova_vis(Y = hand_washing$Bacterial_Counts, 
          group = hand_washing$Method,
          plot_boxplot = FALSE,
          plot_group_variances = FALSE,
          plot_group_sd = FALSE,
          plot_ms_within = FALSE,
          plot_ms_between = FALSE,
          plot_unit_line = TRUE,
          plot_grand_mean = TRUE,
          plot_sd_line = FALSE,
          plot_pooled_sd = FALSE,
          plot_between_group_variances = FALSE,
          ylab = 'Bacterial Counts'
)
```


---
# Within Group Variance (error)

$$SS_{within} = \sum^{}_{k} \sum^{}_{i} (\bar{x}_{ik} -\bar{x}_{k} )^{2}$$
---
# Within Group Variance (error)


```{r, echo=FALSE}
anova_vis(Y = hand_washing$Bacterial_Counts, 
          group = hand_washing$Method,
          plot_boxplot = FALSE,
          plot_group_variances = TRUE,
          plot_group_sd = TRUE,
          plot_ms_within = FALSE,
          plot_ms_between = FALSE,
          plot_unit_line = TRUE,
          plot_grand_mean = TRUE,
          plot_sd_line = FALSE,
          plot_pooled_sd = FALSE,
          plot_between_group_variances = FALSE,
          ylab = 'Bacterial Counts'
)
```


---
# Within Group Variance (error)

```{r, echo=FALSE}
anova_vis(Y = hand_washing$Bacterial_Counts, 
          group = hand_washing$Method,
          plot_boxplot = FALSE,
          plot_group_variances = TRUE,
          plot_group_sd = TRUE,
          plot_ms_within = TRUE,
          plot_ms_between = FALSE,
          plot_unit_line = TRUE,
          plot_grand_mean = TRUE,
          plot_sd_line = FALSE,
          plot_pooled_sd = FALSE,
          plot_between_group_variances = FALSE,
          ylab = 'Bacterial Counts'
)
```


---
# Within Group Variance (error)

```{r, echo=FALSE}
anova_vis(Y = hand_washing$Bacterial_Counts, 
          group = hand_washing$Method,
          plot_boxplot = FALSE,
          plot_group_variances = FALSE,
          plot_group_sd = TRUE,
          plot_ms_within = TRUE,
          plot_ms_between = FALSE,
          plot_unit_line = TRUE,
          plot_grand_mean = TRUE,
          plot_sd_line = FALSE,
          plot_pooled_sd = TRUE,
          plot_between_group_variances = FALSE,
          ylab = 'Bacterial Counts'
)
```

---
# Between Group Variance

$$SS_{between} = \sum^{}_{k} n_{k}(\bar{x}_{k} -\bar{x} )^{2}$$


---
# Between Group Variance

```{r, echo=FALSE}
anova_vis(Y = hand_washing$Bacterial_Counts, 
          group = hand_washing$Method,
          plot_boxplot = FALSE,
          plot_group_variances = FALSE,
          plot_group_sd = TRUE,
          plot_ms_within = FALSE,
          plot_ms_between = FALSE,
          plot_unit_line = TRUE,
          plot_grand_mean = TRUE,
          plot_sd_line = FALSE,
          plot_pooled_sd = FALSE,
          plot_between_group_variances = TRUE,
          ylab = 'Bacterial Counts'
)
```


---
# Between Group Variance

```{r, echo=FALSE}
anova_vis(Y = hand_washing$Bacterial_Counts, 
          group = hand_washing$Method,
          plot_boxplot = FALSE,
          plot_group_variances = FALSE,
          plot_group_sd = TRUE,
          plot_ms_within = FALSE,
          plot_ms_between = TRUE,
          plot_unit_line = TRUE,
          plot_grand_mean = TRUE,
          plot_sd_line = FALSE,
          plot_pooled_sd = FALSE,
          plot_between_group_variances = FALSE,
          ylab = 'Bacterial Counts'
)
```

---
# Mean Square

| Source                  | Sum of Squares                                              | *df*  | MS                                   |
| ------------------------|:-----------------------------------------------------------:|:-----:|:------------------------------------:|
| Between Group (Treatment) | $\sum^{}_{k} n_{k}(\bar{x}_{k} -\bar{x} )^{2}$              | k - 1 | $\frac{SS_{between}}{df_{between}}$ 
| Within Group (Error)    | $\sum^{}_{k} \sum^{}_{i} (\bar{x}_{ik} -\bar{x}_{k} )^{2}$  | n - k | $\frac{SS_{within}}{df_{within}}$    |
| Total                   | $\sum_{n} ({x}_{n} -\bar{x} )^{2}$      | n - 1 |                                      |



---
# $MS_{Between} / MS_{Within}$ = F-Statistic

Mean squares can be represented as squares, hence the ratio of area of the two rectagles is equal to $\frac{MS_{Between}}{MS_{Within}}$ which is the F-statistic.

```{r, echo=FALSE, fig.height=6}
anova_vis(Y = hand_washing$Bacterial_Counts, 
          group = hand_washing$Method,
          plot_boxplot = FALSE,
          plot_group_variances = FALSE,
          plot_group_sd = TRUE,
          plot_ms_within = TRUE,
          plot_ms_between = TRUE,
          plot_unit_line = TRUE,
          plot_grand_mean = TRUE,
          plot_sd_line = FALSE,
          plot_pooled_sd = FALSE,
          plot_between_group_variances = FALSE,
          ylab = 'Bacterial Counts'
)
```


---
# Washing type all the same?

$H_0: \mu_1 = \mu_2 = \mu_3 = \mu_4$

Variance components we need to evaluate the null hypothesis:

* Between Sum of Squares: $SS_{between} = \sum^{}_{k} n_{k}(\bar{x}_{k} -\bar{x} )^{2}$

* Within Sum of Squares: $SS_{within} = \sum^{}_{k} \sum^{}_{i} (\bar{x}_{ik} -\bar{x}_{k} )^{2}$

* Between degrees of freedom: $df_{between} =  k - 1$ (k = number of groups)

* Within degrees of freedom: $df_{within} =  k(n - 1)$

* Mean square between (aka treatment): $MS_{T} = \frac{SS_{between}}{df_{between}}$

* Mean square within (aka error): $MS_{E} = \frac{SS_{within}}{df_{within}}$


```{r, eval=FALSE, echo=FALSE}
( ss_between <- sum(desc$n * (desc$mean - mean(hand_washing$Bacterial_Counts))^2) )
( ss_within <- sum((tmp$Bacterial_Counts - tmp$mean)^2) )
( ss_total <- sum((hand_washing$Bacterial_Counts - mean(hand_washing$Bacterial_Counts))^2) )
tmp <- merge(hand_washing, desc[,c('group1','mean')], by.x = 'Method', by.y = 'group1')
ss_between + ss_within == ss_total # Verify everything adds up
var(hand_washing$Bacterial_Counts) / 4
```


---
# Comparing $MS_T$ (between) and $MS_E$ (within)

.pull-left[
Assume each washing method has the same variance.

Then we can pool them all together to get the pooled variance ${ s }_{ p }^{ 2 }$

Since the sample sizes are all equal, we can average the four variances: ${ s }_{ p }^{ 2 } = `r prettyNum(mean(desc$Var), digits = 6)`$

```{r}
mean(desc$Var)
```
]

--

.pull-right[
$MS_T$

* Estimates ${ s }_{ p }^{ 2 }$ if $H_0$ is true
* Should be larger than ${ s }_{ p }^{ 2 }$ if $H_0$ is false

$MS_E$

* Estimates ${ s }_{ p }^{ 2 }$ whether $H_0$ is true or not
* If $H_0$ is true, both close to ${ s }_{ p }^{ 2 }$, so $MS_T$ is close to $MS_E$

Comparing

* If $H_0$ is true, $\frac{MS_T}{MS_E}$ should be close to 1
* If $H_0$ is false, $\frac{MS_T}{MS_E}$ tends to be > 1
]

---
class: font120
# The F-Distribution 

* How do we tell whether $\frac{MS_T}{MS_E}$ is larger enough to not be due just to random chance?

* $\frac{MS_T}{MS_E}$ follows the F-Distribution
	* Numerator df:  k - 1 (k = number of groups)
	* Denominator df:  k(n - 1)  
	* n = # observations in each group
	
* $F = \frac{MS_T}{MS_E}$ is called the F-Statistic.

A Shiny App by Dr. Dudek to explore the F-Distribution: <a href='https://shiny.rit.albany.edu/stat/fdist/' window='_new'>https://shiny.rit.albany.edu/stat/fdist/</a>

---
# The F-Distribution (cont.) 

```{r fdistribution, fig.width=10, fig.height=6, tidy=FALSE, eval=TRUE}
df.numerator <- 4 - 1
df.denominator <- 4 * (8 - 1)
DATA606::F_plot(df.numerator, df.denominator, cv = qf(0.95, df.numerator, df.denominator))
```


---
class: font120
# ANOVA Table


| Source                  | Sum of Squares                                              | *df*  | MS                                   | F                                   | p                              |
| ------------------------|:-----------------------------------------------------------:|:-----:|:------------------------------------:|:-----------------------------------:|--------------------------------|
| Between Group (Treatment) | $\sum^{}_{k} n_{k}(\bar{x}_{k} -\bar{x} )^{2}$              | k - 1 | $\frac{SS_{between}}{df_{between}}$  | $\frac{MS_{between}}{MS_{within}}$  | area to right of $F_{k-1,n-k}$ |
| Within Group (Error)    | $\sum^{}_{k} \sum^{}_{i} (\bar{x}_{ik} -\bar{x}_{k} )^{2}$  | n - k | $\frac{SS_{within}}{df_{within}}$    |                                     |                                |
| Total                   | $\sum_{n} ({x}_{n} -\bar{x} )^{2}$      | n - 1 |                                      |                                     |                                |


```{r}
aov(Bacterial_Counts ~ Method, data = hand_washing) |> summary()
```


---
# Assumptions and Conditions

* To check the assumptions and conditions for ANOVA, always look at  the side-by-side boxplots.
	* Check for outliers within any group.
	* Check for similar spreads.
	* Look for skewness.
	* Consider re-expressing.
	
* Independence Assumption
	* Groups must be independent of each other.
	* Data within each group must be independent.
	* Randomization Condition
	
* Equal Variance Assumption
	* In ANOVA, we pool the variances.  This requires equal variances from each group:  Similar Spread Condition.


---
# More Information

ANOVA Vignette in the `VisualStats` package: https://jbryer.github.io/VisualStats/articles/anova.html

The plots were created using the `VisualStats::anova_vis()` function.

Shiny app:

```{r, eval=FALSE}
# remotes::install_github('jbryer/VisualStats')
library(VisualStats)
VisualStats::anova_shiny()
```


---
class: font120
# What Next? 

* P-value large -> Nothing left to say
* P-value small -> Which means are large and which means are small?
* We can perform a t-test to compare two of them.
* We assumed the standard deviations are all equal.
* Use $s_p$, for pooled standard deviations.
* Use the Students t-model, df = N - k.
* If we wanted to do a t-test for each pair:
	* P(Type I Error) = 0.05 for each test.
	* Good chance at least one will have a Type I error.
* **Bonferroni to the rescue!**
	* Adjust a to $\alpha/J$ where J is the number of comparisons.
	* 95% confidence (1 - 0.05) with 3 comparisons adjusts to $(1 - 0.05/3) \approx  0.98333$.
	* Use this adjusted value to find t**.

---
# Multiple Comparisons (no Bonferroni adjustment)

.code80[
```{r, fig.height=5, tidy=FALSE}
cv <- qt(0.05, df = 15)
tab <- describeBy(hand_washing$Bacterial_Counts, group = hand_washing$Method, mat = TRUE)
ggplot(hand_washing, aes(x = Method, y = Bacterial_Counts)) + geom_boxplot() + 
	geom_errorbar(data = tab, aes(x = group1, y = mean, 
								  ymin = mean - cv * se, ymax = mean + cv * se), 
				  color = 'darkgreen', width = 0.5, size = 1) +
	geom_point(data = tab, aes(x = group1, y = mean), color = 'blue', size = 3)
```
]

---
# Multiple Comparisons (3 paired tests)

.code80[
```{r, fig.height=5, tidy=FALSE}
cv <- qt(0.05 / 3, df = 15)
tab <- describeBy(hand_washing$Bacterial_Counts, group = hand_washing$Method, mat = TRUE)
ggplot(hand_washing, aes(x = Method, y = Bacterial_Counts)) + geom_boxplot() + 
	geom_errorbar(data = tab, aes(x = group1, y = mean, 
								  ymin = mean - cv * se, ymax = mean + cv * se), 
				  color = 'darkgreen', width = 0.5, size = 1) +
	geom_point(data = tab, aes(x = group1, y = mean), color = 'blue', size = 3)
```
]

---
# Multiple Comparisons (6 paired tests)

.code80[
```{r, fig.height=5, tidy=FALSE}
cv <- qt(0.05 / choose(4, 2), df = 15)
tab <- describeBy(hand_washing$Bacterial_Counts, group = hand_washing$Method, mat = TRUE)
ggplot(hand_washing, aes(x = Method, y = Bacterial_Counts)) + geom_boxplot() + 
	geom_errorbar(data = tab, aes(x = group1, y = mean, 
								  ymin = mean - cv * se, ymax = mean + cv * se ), 
				  color = 'darkgreen', width = 0.5, size = 1) +
	geom_point(data = tab, aes(x = group1, y = mean), color = 'blue', size = 3)
```
]



---
class: left, font140
# One Minute Paper

.pull-left[
1. What was the most important thing you learned during this class?
2. What important question remains unanswered for you?
]
.pull-right[
```{r, echo=FALSE, fig.width=5, fig.height=5}
qrcode::qr_code(one_minute_paper) |> plot(col = c('#FAFAFA', 'black'))
```
]

`r one_minute_paper`
