---
title: "R Coding Basics"
name: "YOUR NAME"
output:
  pdf_document: default
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
library(tidyverse)
library(kableExtra)
```


1.  Why does this code not work?

```{r}
#| error: true

my_variable <- 10
my_varıable
```

Look carefully! (This may seem like an exercise in pointlessness, but training your brain to notice even the tiniest difference will pay off when programming.)


2.  Tweak each of the following R commands so that they run correctly:

```{r}
#| eval: false

libary(todyverse)

ggplot(dTA = mpg) + 
  geom_point(maping = aes(x = displ y = hwy)) +
  geom_smooth(method = "lm)
```



3.  Press Option + Shift + K / Alt + Shift + K. What happens? How can you get to the same place using the menus?




4.  Run the following lines of code. Which of the two plots is saved as `mpg-plot.png`?  Why?

```{r}
#| eval: false

my_bar_plot <- ggplot(mpg, aes(x = class)) +
  geom_bar()
my_scatter_plot <- ggplot(mpg, aes(x = cty, y = hwy)) +
  geom_point()
ggsave(filename = "mpg-plot.png", plot = my_bar_plot)
```

